---
title: "Exploring Data"
author: "Sara Stoudt"
date: "1/15/2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE,message=FALSE}
library(learnr)
library(tidyverse)
library(gradethis)
library(openintro)
library(testthat)
tutorial_options(
  # code running in exercise times out after 30 seconds
  exercise.timelimit = 30, 
  # use gradethis for checking
  exercise.checker = gradethis::grade_learnr
  )
knitr::opts_chunk$set(echo = FALSE)

```

## Let's Explore

```{r load-data, exercise = T}
data(arbuthnot)
```

Let's start to examine the data a little more closely. We can access the data in
a single column of a data frame separately.

```{r view-boys, exercise = T}
arbuthnot$boys
```

This command will only show the number of boys baptized each year. 
The dollar sign basically says "go to the data frame that comes before me, and find the variable that comes after me".

What command would you use to extract just the counts of girls baptized? Try
   it!
   
```{r view-girls, exercise = T}
___$___
```

```{r view-girls-solution}
arbuthnot$girls
```

    
Notice that the way R has printed these data is different. When we looked at the complete data frame, we saw 82 rows, one on each line of the display. 
These data are no longer structured in a table with other variables, so they are displayed  one right after another. 
Objects that print out in this way are called *vectors*; they represent a set of numbers. 
R has added numbers in [brackets] along the left side of the printout to indicate locations within the vector. For example, 5218  follows `[1]`, indicating that `5218` is the first entry in the vector. 
And if `[43]` starts a line, then that would mean the first number on that line would represent the 43rd entry in the vector.


## Data visualization

R has some powerful functions for making graphics. We can create a simple plot of the number of girls baptized per year with the command

```{r plot-girls-vs-year, exercise = T}
ggplot(data = arbuthnot, aes(x = year, y = girls)) + 
  geom_point()
```


We use the `ggplot()` function to build plots. 
If you run the plotting code in your console, you should see the plot appear under the *Plots* tab of the lower right panel of RStudio. If you run the chunk, the plot will appear beneath the code chunk.
Notice that the command above again looks like a function, this time with arguments separated by commas. 

With `ggplot()`:

- The first argument is always the dataset. 
- Next, you provide the variables from the dataset to be assigned to `aes`thetic elements of the plot, e.g. the x and the y axes. 
- Finally, you use another layer, separated by a `+` to specify the `geom`etric object for the plot. Since we want to scatterplot, we use `geom_point()`.

For instance, if you wanted to visualize the above plot using a line graph, you would replace `geom_point()` with `geom_line()`.

```{r plot-girls-vs-year-line, exercise = T}
ggplot(data = arbuthnot, aes(x = year, y = girls)) + 
  geom_line()
```

You might wonder how you are supposed to know the syntax for the `ggplot` function. 
Thankfully, R documents all of its functions extensively. 
To learn what a function does and its arguments that are available to you, use the "Help" tab in the bottom right.


How would you create a plot that would help you answer the following question:
Is there an apparent trend in the number of boys baptized over the years? 

                              
```{r plot-boys-vs-year-point, exercise = T}
ggplot(data = ___, aes(x = ___, y = ___)) + 
  geom_point()
```


```{r plot-boys-vs-year-point-solution}
ggplot(data = arbuthnot, aes(x =year, y = boys)) + 
  geom_point()
```

What if you wanted a line instead of points?

```{r plot-boys-vs-year-line, exercise = T}
ggplot(data = ___, aes(x = ___, y = ___)) + ___
```


```{r plot-boys-vs-year-line-solution}
ggplot(data = arbuthnot, aes(x =year, y = boys)) + 
  geom_line()
```

## R as a big calculator

Now, suppose we want to plot the total number of baptisms. 
To compute this, we could use the fact that R is really just a big calculator. 
We can type in mathematical expressions like

```{r calc-total-bapt-numbers, exercise = T}
5218 + 4683
```

to see the total number of baptisms in 1629. We could repeat this once for each year, but there is a faster way. 
If we add the vector for baptisms for boys to that of girls, R will compute all sums simultaneously.

```{r calc-total-bapt-vars, exercise = T}
arbuthnot$boys + arbuthnot$girls
```

What you will see are 82 numbers (in that packed display, because we aren’t looking at a data frame here), each one representing the sum we’re after. 
Take a look at a few of them and verify that they are right.

## Adding a new variable to the data frame

We'll be using this new vector to generate some plots, so we'll want to save it as a permanent column in our data frame.

```{r calc-total-bapt-vars-save, exercise = T}
arbuthnot <- arbuthnot %>%
  mutate(total = boys + girls)
```

The `%>%` operator is called the **piping** operator. It takes the output of the previous expression and pipes it into the first argument of the function in the following one. 
To continue our analogy with mathematical functions, `x %>% f(y)` is equivalent to `f(x, y)`.


**A note on piping:** Note that we can read these two lines of code as the following: *"Take the `arbuthnot` dataset and **pipe** it into the `mutate` function. 
Mutate the `arbuthnot` data set by creating a new variable called `total` that is the sum of the variables called `boys` and `girls`. 

Then assign the resulting dataset to the object called `arbuthnot`, i.e. overwrite the old `arbuthnot` dataset with the new one containing the new variable."*
  
This is equivalent to going through each row and adding up the `boys` and `girls` counts for that year and recording that value in a new column called `total`.


**Where is the new variable?** When you make changes to variables in your dataset, click on the name of the dataset again to update it in the data viewer.

  
You'll see that there is now a new column called `total` that has been tacked onto the data frame.

The special symbol `<-` performs an *assignment*, taking the output of one line of code and saving it into an object in your environment. 
In this case, you already have an object called `arbuthnot`, so this command updates that data set with the new mutated column.

You can make a line plot of the total number of baptisms per year with the command

```{r plot-total-vs-year, exercise = T}
ggplot(data = arbuthnot, aes(x = year, y = total)) + 
  geom_line()
```

Just as you computed the total number of births, you can compute the ratio of the number of boys to the number of girls baptized in 1629 with

```{r calc-prop-boys-to-girls-numbers, exercise = T}
5218 / 4683
```

or you can act on the complete columns with the expression

```{r calc-prop-boys-to-girls-vars, exercise = T}
arbuthnot <- arbuthnot %>%
  mutate(boy_to_girl_ratio = boys / girls)
```

You can also compute the proportion of newborns that are boys in 1629

```{r calc-prop-boys-numbers, exercise = T}
5218 / (5218 + 4683)
```

or you can compute this for all years simultaneously and append it to the dataset

```{r calc-prop-boys-vars, exercise = T}
arbuthnot <- arbuthnot %>%
  mutate(boy_ratio = boys / total)
```

Note that we are using the new `total` variable we created earlier in our calculations.

Generate a plot of the proportion of boys born over time. What do you see? 

```{r plot-prop-boys, exercise = T}
ggplot(data = ___, aes(x = ___, y = ___)) + 
  geom_line()
```

```{r plot-prop-boys-solution}
arbuthnot <- arbuthnot %>% mutate(total = boys + girls) %>%
  mutate(boy_ratio = boys / total)

ggplot(data = arbuthnot, aes(x =year, y = boy_ratio)) + 
  geom_line()
```

Finally, in addition to simple mathematical operators like subtraction and division, you can ask R to make comparisons like greater than, `>`, less than, `<`, and equality, `==`. 
For example, we can ask if the number of births of boys outnumber that of girls in each year with the expression

```{r boys-more-than-girls, exercise = T}
arbuthnot <- arbuthnot %>%
  mutate(more_boys = boys > girls)
```

This command adds a new variable to the `arbuthnot` dataframe containing the values of either `TRUE` if that year had more boys than girls, or `FALSE` if that year did not (the answer may surprise you). 
This variable contains a different kind of data than we have encountered so far. 
All other columns in the `arbuthnot` data frame have values that are numerical (the year, the number of boys and girls). 
Here, we've asked R to create *logical* data, data where the values are either `TRUE` or `FALSE`. 
In general, data analysis will involve many different kinds of data types, and one reason for using R is that it is able to represent and compute with many of them.

## More Practice

In the previous few exercises, you recreated some of the displays and preliminary analysis of Arbuthnot's baptism data. 
Your assignment involves repeating these steps, but for present day birth records in the United States. 
The data are stored in a data frame called `present`. These data come from reports by the Centers for Disease Control. 
You can learn more about them by bringing up the help file using the command `?present`.

To find the minimum and maximum values of columns, you can use the functions `min` and `max` within a `summarise()` call, which you will learn more about in the following lab. 
Here's an example of how to find the minimum and maximum amount of boy births in a year for the `arbuthnot` data:
  
```{r summarize_min_and_max, exercise = T}
arbuthnot %>%
  summarise(min = min(boys), max = max(boys))
```

What years are included in the `present` data set? 

```{r range, exercise = T}
___ %>%
  summarise(min = min(___), max = max(___))
```

```{r range-solution}
present %>%
  summarise(min = min(year), max = max(year))
```


What are the dimensions of the data frame? 

```{r dim, exercise = T}
glimpse(___)
```

You want it to look like this.

```{r dim-solution}
glimpse(present)
```

What are the variable (column) names?

```{r names, exercise = T}
names(___)
```

You want it to look like this.

```{r names-solution}
names(present)
```
  
How do these counts compare to Arbuthnot's? 

```{r compare, exercise = T}
arbuthnot %>%
  summarise(min = min(boys), max = max(boys))

___ %>% 
   summarise(min = min(boys), max = max(boys))
```

```{r compare-solution}
arbuthnot %>%
  summarise(min = min(boys), max = max(boys))

present %>% 
   summarise(min = min(boys), max = max(boys))
```


```{r magnitude}
question("Are they of a similar magnitude?",
    answer("Yes"),
    answer("No, Arbuthnot's are greater."),
    answer("No, the US's are greater.", correct = T),
    allow_retry = TRUE,
    random_answer_order = F
  )
```


Make a plot that displays the proportion of boys born over time. 

```{r plot-prop-boys-2, exercise = T}
present <- present %>% mutate(boy_ratio = ___)

ggplot(data = ___, aes(x = ___, y = ___)) + 
  geom_line()
```


```{r plot-prop-boys-2-solution}
present <- present %>% mutate(boy_ratio = boys/(boys+girls))

ggplot(data = present, aes(x = year, y = boy_ratio)) + 
  geom_line()
```


```{r obs-hold}
question("Does Arbuthnot's observation about boys being born in greater proportion than girls hold up in the U.S.?",
    answer("Yes",correct = T),
    answer("No"),
    allow_retry = TRUE,
    random_answer_order = F
  )
```




In what year did we see the most total number of births in the U.S.? *Hint:*
First calculate the totals and save it as a new variable. Then, sort your
dataset in descending order based on the total column. We can arrange the data 
in a descending order with another function: `desc` (for descending order).


```{r sample-arrange, exercise = T}
present <- present %>%
  mutate(total = ___)

present %>%
  arrange(desc(total))
```



```{r sample-arrange-solution}
present <- present %>%
  mutate(total = boys + girls)

present %>%
  arrange(desc(total))
```

